Dependent Object Types

### 2012-06-30 doc/semester/report.tex
non-expanding types and subtyping transitivity

\subsection{Subtyping Transitivity and Preservation}

Some subtyping transitivity is essential for soundness. Indeed, I show
how to construct a counterexample to preservation from any
counterexample to subtyping transitivity where the three involved
types $S, T, U$ are expressible within a universe $u$ but not
necessarily realizable:

@@

Consider the simplest possible type that refers to itself: $T = \Top
\refine z {\Ldecl L \Bot z.L}$. This type is well-formed but
non-expanding.

Non-expanding types are problematic for subtyping transitivity.

>>>>

\subsection{Subtyping Transitivity and Preservation}\label{subpres}

Some subtyping transitivity is essential for soundness. Indeed, I show
how to construct a counterexample to preservation from any
counterexample to subtyping transitivity where the three involved
types $S, T, U$ are expressible within a realizable universe, though
the types themselves don't need to be realizable:

@@

Consider the simplest possible type that refers to itself: $p \typ T =
\Top \refine z {\Ldecl L \Bot z.L}$. Now, $p.L$ is well-formed but
non-expanding. Indeed, the only expansion rule that applies to $p.L$
is $\textsc{tsel-$\expand$}$, but it requires expanding the upper
bound... $p.L$! Thus, there is no finite derivation for the expansion
of $p.L$.

Non-expanding types are problematic for subtyping
transitivity. Consider: $p.L \sub \Top$ by \textsc{$\sub$-$\Top$} and
$\Top \sub \Top \refine z {}$ by \textsc{$\sub$-rfn}. By transitivity
on $\Top$, we expect $p.L \sub \Top \refine z {}$, but we cannot infer
this from the rules. The only rules that apply are \textsc{$\sub$-rfn}
and \textsc{tsel-$\sub$}. \textsc{$\sub$-rfn} gets stuck expanding
$p.L$. \textsc{tsel-$\sub$} gets stuck because the conclusion is also
a premise as $p \ni \Ldecl L \Bot p.L$.

For this counterexample, it seems enough to just add transitivity on
$\Top$ as a built-in rule. But this doesn't solve the general issue
that non-expanding types don't play well with subtyping. The problem
is deep, as it's possible to construct a realizable universe with
three non-trivial types that fail subtyping transitivity.

Consider an environment where $u$ is bound to:
\begin{align*}
\Top & \mlrefine u {\\
&\ \Ldecl {\mathit{Bad}} {\Bot} {u.\mathit{Bad}}\\
&\ \Ldecl {\mathit{Good}} {\Top \refine z {\Ldecl L \Bot \Top}} {\Top \refine z {\Ldecl L \Bot \Top}}\\
&\ \Ldecl {\mathit{Lower}} {u.\mathit{Bad} \tand u.\mathit{Good}} {u.\mathit{Good}}\\
&\ \Ldecl {\mathit{Upper}} {u.\mathit{Good}} {u.\mathit{Bad} \tor u.\mathit{Good}}\\
&\ \Ldecl X {u.\mathit{Lower}} {u.\mathit{Upper}}\\
}&
\end{align*}
Note that each lower bound is a subtype of its upper bound, so $u$ is
realizable. Hence, if we find a counterexample to subtyping
transitivity with types expressible within $u$, we can apply the trick
from section~\ref{subpres} to create a counterexample to preservation.

Indeed, here is such a counterexample to subtyping transitivity:
\begin{align*}
S &= u.\mathit{Bad} \tand u.\mathit{Good}\\
T &= u.\mathit{Lower}\\
U &= u.X \refine z {\Ldecl L \Bot \Top}
\end{align*}
We have $S <: T$ and $T <: U$, but we cannot derive $S <: U$ because
$S$ doesn't expand.
